\section{Limitations --- draft v1}
\label{sec:limitations}

Stated as facts about the artifact, each with its consequence.

1. \textbf{The curated fallback exists.} Diagnostics constructed through
   an engine with no recorder attached carry hand-written reference
   chains, not recorded ones. In the compiled pipeline this path is
   unreachable (verified by exhaustive search: three engines, three
   constructor-attached recorders, no other instantiation), but the
   dual source of truth remains a maintenance hazard bounded only by
   tests.
2. \textbf{Scope placement is selective.} Recorded traces are coarse to
   the granularity of the placement rule (D9); omission cannot
   fabricate a path, but a frame the rule skips is a frame the user
   never sees.
3. \textbf{Per-instruction provenance is not transitive} under dead-code
   elimination (\S6.4): a folded instruction's history dies with it
   once propagation makes it dead. Locked in by tests as documented
   semantics; a dataflow-chained design is future work.
4. \textbf{Five diagnostic kinds are reserved-unreachable} --- defined,
   fully worded, never emitted, because the language cannot yet
   produce them. Tests are forbidden from expecting them; counting
   them as supported would inflate the taxonomy by 29\%.
5. \textbf{The cascade invariants are shape-specific.} Mutation testing
   measured the residual: 3\% of single-fault mutants still produce
   three or more diagnostics (\S9.3).
6. \textbf{Assembly provenance is comments, not debug info.} Readable by
   humans, invisible to debuggers; no DWARF/\texttt{.loc} emission.
7. **The knowledge base centralizes prose but implements no
   internationalization** --- centralization makes i18n \emph{possible}
   (one place to translate), and we did none of it.
8. \textbf{Single translation unit, no calls, no control flow.} The
   cascade and trace-depth results inherit the language's flatness;
   \S10.2 discusses what deeper languages would demand.
9. \textbf{The recorder is single-threaded by construction} --- one
   recorder per stage, no synchronization; a parallel front end
   would need per-thread recorders and a merge discipline.
10. \textbf{Windows-locked build and measurement.} The Makefile's shell
    commands, the harness, and all numbers are from one platform
    (\S8.4, \S11).



